The evolution of gene expression in the term placenta of viviparous mammals

### 2016-02-12 manuscript/manuscript.Rnw
add eviatar nevo to authors list
add derek changes to introduction

\begin{frontmatter}
\begin{fmbox}
\dochead{Research}
\title{The evolution of gene expression in the term placenta of viviparous mammals}%
%
\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
\author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL} \fnm{Amy} \snm{Weckle}}
\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}

@@

  \parttitle{Conclusions} While there are multiple non-housekeeping
  genes which are highly expressed in all placentas examined and are
  likely to be involved in the basic functions of the mammalian
  placenta, we see incredible lineage-specific diversity in the
  expression and expansion of many genes. These lineage-specific genes
  are likely to play a role in the anatomical variability of the
  placenta and may also be implicated in important \human placental
  disorders such as pre-eclampsia.

@@

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental”
mammals. The term “placental mammals” is misleading, as marsupials
also possess a placenta, although its direct apposition to maternal
tissues lasts for a proportionally shorter time during gestation
\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. Marsupials also
differ from eutherian mammals in that most possess only a
choriovitelline placenta, produced from the fusion of the chorion and
yolk sac, during most of gestation. Although some eutherian mammals
may possess a choriovitelline placenta at some point in development,
the major site of nutrient exchange in placentals occurs via the
chorioallantoic placenta, produced from the fusion of the chorion and
allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
whether the first placenta was more like that of marsupials or
eutherians; however, using phylogenetic methods, the ancestral
eutherian was reconstructed as possessing a relatively invasive
hemochorial placenta, in which the maternal and fetal blood supply is
separated by one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \human, \mouse, \cow) have
revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\carmeli, \galili, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \elephant, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{P. troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

One symptom of this maternal-fetal conflict is its potential affect on
\human health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, \human possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of pre-eclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in \textit{P. troglodytes}
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
\textit{Gorilla gorilla}
\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
far higher incidence in \human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
  troglodytes} and \textit{G. gorilla} also have deeply invasive
trophoblasts, although the progression of invasion in great apes may
be slower than \human
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \bonobo (the sister
species to \textit{P. troglodytes}), \smonkey, \mouse, \galili,
\carmeli, \dog, \cow, \elephant, \armadillo, and \opossum. These
species represent three major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria) as well as marsupials.
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation (\cref{tab:gestational_periods}). We use this data, along with
existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify \human specific differences in gene
expression that may be related to obstetrical disorders such as
pre-eclampsia.

@@

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled 
using Trinity v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence}
and annotated using Diamond
\cite{Buchfink.ea2015:Fastandsensitiveprotein} against either the
\human or \mouse transcriptome
(Supplemental \cref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \bonobo. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

@@

All 1:1 orthologues present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart
\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
package
\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
analyses of gene expression evolution described below.

>>>>

\begin{frontmatter}
\begin{fmbox}
\dochead{Research}
\title{The evolution of gene expression in the term placenta of viviparous mammals}%
%
\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
\author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL} \fnm{Amy} \snm{Weckle}}
\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}

@@

  \parttitle{Conclusions} While there are multiple non-housekeeping
  genes which are highly expressed in all placentas examined and are
  likely to be involved in the basic functions of the mammalian
  placenta, we see incredible lineage-specific diversity in the
  expression and expansion of many genes. These lineage-specific genes
  are likely to play a role in the anatomical variability of the
  placenta and may also be implicated in important human placental
  disorders such as pre-eclampsia.

@@

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental
mammals”. The term “placental mammals” is misleading, as extant
yinothereians (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, in which the maternal and fetal blood supply is separated by
one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
\DLA{To discuss with derek}

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Nanospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

One symptom of this maternal-fetal conflict is its potential affect on
health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, \human possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of pre-eclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in \textit{P. troglodytes}
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
\textit{Gorilla gorilla}
\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
far higher incidence in \human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
  troglodytes} and \textit{G. gorilla} also have deeply invasive
trophoblasts, although the progression of invasion in great apes may
be slower than \human
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \textit{Pan paniscus}
(the sister species to \textit{P. troglodytes}), \textit{Ateles
  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
\cow, \elephant, \textit{Dasypus novemcinctus}, and
\textit{Monodelphis domesticus}. These species represent four major
branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
wide array of placental morphologies as well as traits such as number
and size of offspring and length of gestation
(\cref{tab:gestational_periods}). We use this data, along with
existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify \human specific differences in gene
expression that may be related to obstetrical disorders such as
pre-eclampsia.

@@

\DLA{Add QC information}

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled 
using Trinity v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence}
and annotated using Diamond
\cite{Buchfink.ea2015:Fastandsensitiveprotein} against either the
\human or \mouse transcriptome
(Supplemental \cref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{P. troglodytes} was used as
a reference for \bonobo. The separation between the two recognized Pan
species is generally perceived to be $\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

@@

All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart
\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
package
\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
analyses of gene expression evolution described below.
